---
title: "R-exercise 5"
author: "Yan Pan"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plotly)
library(DT)
```

# Exercise 1 (Muscle Strength)

```{r}
library(nlme)
wlong <- read.table("http://users.jyu.fi/~slahola/files/lda_datoja/wlong.txt", 
                    header = TRUE)
wlong$trt <- factor(wlong$trt)

weights.lme <- lme(st ~ time * trt, random = ~ 1 + time | id, data = wlong)
summary(weights.lme)$coefficients$fixed
``` 

Conclusions based on fixed effect: trt2 (increasing weights) seems to be positively beneficial to muscle strength.

```{r}
weights.2 <- lme(st ~ time * trt, random = ~ 1 |id, data = wlong)
anova(weights.lme, weights.2)
``` 

Model comparison suggest the original model is better, i.e. the random effect multiplier is indeed necessary.

```{r}
# summary(weights.lme)$coefficients also gives a nice table
datatable(data.frame(random.effects(weights.lme)))
```

```{r results='hide'}
plotdata <- cbind(wlong, weights.lme$fitted)
plot1 <- ggplot(aes(x = time, y = fixed, color = trt), data = plotdata) + 
  geom_line() + 
  geom_point(aes(y = st)) +
  facet_wrap(~id) +
  ggtitle("Lines are fitted values, points are the observed")
```

```{r}
ggplotly(plot1)
``` 

# Exercise 2 (PISA-maths)
```{r}
pisa <- read.table("http://users.jyu.fi/~slahola/files/lda_datoja/pisamat.txt",
                   header = TRUE)
plot(mathscore ~ factor(SCHOOLID), data = pisa)
``` 

There seems no difference among schools.

```{r}
pisa$alue <- factor(pisa$alue)
pisa$mot <- factor(pisa$mot)
pisa.lm <- lm(mathscore ~ matem + ESCS + alue + mot, data = pisa)
# using summary instead of coef() to get p value also
datatable(data.frame(summary(pisa.lm)[4]))
``` 

Last mathematics schools, social economics status and motivations are somewhat positively influence PISA-mathematics socre.

```{r}
index50 <- which(pisa$SCHOOLID < unique(pisa$SCHOOLID)[51])
plotdata <- data.frame(residuals = pisa.lm$residuals[index50],
                       SchoolID = pisa$SCHOOLID[index50])
plot2 <- ggplot(aes(x=SchoolID, y=residuals), data = plotdata) +
  geom_point() + ggtitle("residual of 50 schools (lm)")
ggplotly(plot2)
``` 

```{r}
pisa.lme <- lme(mathscore ~ matem + ESCS + alue + mot,
                random = ~ 1 | SCHOOLID ,data = pisa)
summary(pisa.lme)$coefficients

sigma <- colMeans(pisa.lme$residuals^2)
sigma
rho <- sigma[2] / sum(sigma)
rho
```

The variance components and correlation are calculated from residuals.

```{r}
plotdata <- data.frame(fitted = pisa.lme$fitted[,2],
                       residuals = pisa.lme$residuals[,2],
                       alue = pisa$alue)
plot3 <- ggplot(aes(x=fitted, y=residuals), data = plotdata) + 
  geom_point() + facet_wrap(~alue) + ggtitle("Residuals of Random Effect")
ggplotly(plot3)
``` 

For different regions, there seem to be concentration difference in residual/fitted.

# Exercise 3 (multi-level model)

```{r}
for (i in 1:nrow(pisa)) {
  index <- which(pisa$SCHOOLID == pisa$SCHOOLID[i])
  pisa$meanmath[i] <- as.numeric(mean(pisa$matem[index]) > 7.6)
  pisa$meanmot[i] <- as.numeric(mean(as.numeric(pisa$mot[index])) > 0.6)
}

pisa.multi <- lme(mathscore ~ matem * meanmath,
                  random = ~ (meanmath + meanmot) | SCHOOLID,
                  data = pisa)
summary(pisa.multi)
``` 