# Sekamallit ja Pitkittäisaineistonjen Analyysi (Demo 2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
load("lda_0.RData")
```

## 1.R pacakge download
```{r results='hide'}
library(plotly)
``` 

## 2.Graphs for longitudinal data
```{r}
names(lyijy) <- c('id', "kas", "vk0", "vk1", "vk4", "vk6")
lyijy.long <- reshape(lyijy,
                      varying = list(c("vk0", "vk1", "vk4", "vk6")),
                      v.names = "lyijypitoisuus",
                      direction = "long")
g2 <- ggplot(aes(x = time, y = lyijypitoisuus, color = kas, group = id), 
              data = lyijy.long) 
g2b <- g2 + geom_point() +  geom_line() + ggtitle("Spagetti Graph")
ggplotly(g2b)

g2c <- g2b + facet_wrap(~kas) + ggtitle ("Spagetti Graph by Treatment")
ggplotly(g2c)
```

By this graph, it is quite obvious that the medicine is effective reducing the lead poisoning on week 1 through 6.

```{r}
g2d <- ggplot(aes(x = time, y = lyijypitoisuus, color = kas), data = lyijy.long) + 
  geom_point() + 
  facet_wrap(~kas) +
  geom_smooth(method = "loess") +
  ggtitle("comparison graph with LOESS smoothing")

ggplotly(g2d)

pickid <- sample(1:100, 20)
lyijy.20 <- lyijy.long[which(lyijy.long$id %in% pickid),]
g2e <- ggplot(aes(x = time, y = lyijypitoisuus, color = kas), data = lyijy.20) +
  geom_point() + 
  facet_wrap(~kas) +
  geom_smooth(method = "loess") +
  ggtitle("comparison of 20 random samples")

ggplotly(g2e)
``` 


## Exercise 3
```{r}

g3 <- ggplot(aes(x = time, y = lyijypitoisuus, color = kas), data = lyijy.long) 
g3a <- g3 + geom_boxplot() + ggtitle("Boxplot for two treatment groups")
print("Plotly does not give the same plot as native GGplot, but seems to be better?")
ggplotly(g3a)


groupID <- which(lyijy.long$kas == "A")
sample.mean <- c (mean(lyijy.long$lyijypitoisuus[groupID]),
                  mean(lyijy.long$lyijypitoisuus[-groupID]))
sample.sd <- sqrt (c(var(lyijy.long$lyijypitoisuus[groupID]),
                     var(lyijy.long$lyijypitoisuus[-groupID])))
result <- rbind(sample.mean, sample.sd)
colnames(result) <- c("Medicine", "Placebo")
upper <- result[1,] + 1.96 * result[2,]
lower <- result[1,] - 1.96 * result[2,]
rbind(result, upper, lower)
``` 

For the visulization.of correlations, GGPAIRS is used. It requires extra library, but very handy to use. Indeed, the correlation can be seen direclty from the graph.

```{r}
library(GGally)
g3c <- ggpairs(lyijy[,2:6], aes(color = kas, alpha = 0.4))
ggplotly(g3c)
cor(lyijy[,3:6])
``` 

GGplot supports linear model smoothing, and plots directly to the graph. However, I find it easier to get coefficients and residuals using the "lm" function

```{r}
g3d <- g3 + geom_point() + geom_smooth(method = "lm")
ggplotly(g3d)
# "traditional" methods
fit <- lm(lyijypitoisuus ~ kas + time, data = lyijy.long)
lyijy.long$residual <- fit$residuals
g3e <-ggplot(aes(x = 1:400, y = residual, color = kas), data = lyijy.long) + 
  geom_point() + ggtitle("residuals from linear model (showing treatment by color)")
ggplotly(g3e)

``` 

```{r}
attach(lyijy.long)
resData <- data.frame(vk0Res = residual[which(time==1)],
                      vk1Res = residual[which(time==2)],
                      vk4Res = residual[which(time==3)],
                      vk6Res = residual[which(time==4)])
ggplotly(ggpairs(resData) + ggtitle("Residuals Correlations"))
``` 
