---
title: "R-exercise 6"
author: "Yan Pan"
output:
  html_document:
    toc: true
    df_print: paged
    theme: 'readable'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE, comment = "")
library(plotly)

myprint <- function(printD, dig = 3){
  data.frame(round(printD, digits = dig)) 
}
```
##Exercise 1: Generalized Logistic Model
```{r eval = TRUE}
inf <- read.table("http://users.jyu.fi/~slahola/files/lda_datoja/infection.txt")
names(inf) <- c("id", "inf", "trt", "visit")

library(geepack)
inf.gee <- geese(inf ~ trt * visit - trt, id = id, data = inf, 
                 family = binomial, corstr = "exch", scale.fix = TRUE)
betas <- data.frame(summary(inf.gee)$mean)
betas

odds <- data.frame(GroupA.OddsRatio = exp(inf.gee$beta[2]),
                   GroupB.oddsRatio = exp(inf.gee$beta[2] + inf.gee$beta[3]))
odds
``` 

Although the model yields different values of odds ratios, the P-value of <code>trt*visit</code> item implies statistical non-significance. The treatment does not seem to effective.

```{r eval = TRUE}
betas$upp <- betas$estimate + 1.96 * betas$san.se
betas$lwr <- betas$estimate - 1.96 * betas$san.se
odds["upp",] <- exp(c(betas$upp[2], betas$upp[2]+betas$upp[3]))
odds["lwr",] <- exp(c(betas$lwr[2], betas$lwr[2]+betas$lwr[3]))
odds
``` 

```{r eval = TRUE}
library(lme4)
inf.glmm <- glmer(inf ~ trt * visit - trt + (1|id), 
                  family = binomial, data = inf)
summary(inf.glmm)
``` 

In this model, when consider the <code>inf</code> group, the treatment becomes statistically significant. In reality, this model gives more sense, since it consider the important background inforamtion.

##Exercise 2: Poisson Mixed MOdel

```{r eval = TRUE}
skin <- read.table("http://users.jyu.fi/~slahola/files/lda_datoja/skincan.txt", header = TRUE)
can.glmm <- glmer(cancer ~ year * trt - trt + (1|id), family = poisson, data = skin)
betas <- summary(can.glmm)$coefficients
myprint(betas)
``` 

Based on the P value and the estimates, the treatment is effective.

```{r eval = TRUE}
betas <- cbind(betas, confint(can.glmm)[2:4,])

betas.new <- data.frame(lwr = betas[,5],
                        est = betas[,1],
                        upp = betas[,6])
betas.new["Placebo",] <- betas.new[2,]
betas.new["Treatment",] <- betas.new[2,] + betas.new[3,]

odds <- exp(t(betas.new[4:5,]))
colnames(odds) <- c("Placebo.RiskRatio","Treatment.RiskRatio")
odds

par(mfrow = c(1,3))
plot(skin$age, resid(can.glmm))
plot(skin$exp, resid(can.glmm))
plot(skin$skin, resid(can.glmm))
``` 

The residual plot indicates there are potentially correlation between age and the risk ratio.

```{r eval = TRUE}
can.glmm2 <- glmer(cancer ~ year * trt - trt + age + skin + exp +  (1|id), 
                  family = poisson, data = skin)
myprint(summary(can.glmm2)$coefficients)
``` 

It seems <code>age</code>, <code>skin</code> and <code>exp</code> are all statistically significant.

##Exercise 3: Transitiomallit

```{r eval = TRUE}
depr <- read.table("http://users.jyu.fi/~slahola/files/lda_datoja/tunteet.txt")
names(depr) <- c("id", "terapiamuoto", "sp", "terapiavko", "tunteet")

# "lazy way" to get the lag
# move one row below, but change always the first one to 0
depr$tunteetlag[2:nrow(depr)] <- depr$tunteet
depr$tunteetlag[1] <- 0
for (i in 2:nrow(depr)) {
  if(depr$id[i] != depr$id[i-1]){
    depr$tunteetlag[i] <- 0
  }
}

tunteet.log <- glm(tunteet ~ terapiamuoto + sp + terapiavko + tunteetlag,
                   family = binomial, data = depr)
summary(tunteet.log)
``` 