# TILS644 Monimuuttujamenetelmät (Demo 4)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
load("mmm_0.RData")
```

For simple typesetting, some matrix symbol are not in bold

## exercise 1

$\begin{align} cov(\boldsymbol{X} )&=  E(X-\mu)(X-\mu)'  \\ &=  EXX' & \text{since } X \text{ is standard} \\ &=  E(Af+e)(Af+e)' \\ &=  cov(Af+e) \\ &=  cov(Af) + cov(e) & \text{becasue of incorrelation} \\ &=  Acov(f) A' + cov(e) & \text{affine transformation} \\ &= \boldsymbol{ A\Phi A'}  + \boldsymbol\Psi  \end{align}$ 

<br>

## exercise 2

$\begin{align} cov(\boldsymbol{X,f} )&= E(X-\mu_X)(f-\mu_f)' \\ &= E[Xf'] \\ &=  E[(Af+e)f'] \\ &= E[Aff' + ef'] \\ &=  AE[ff'] + E[ef'] & \text{linearity of expectation} \\ &=  A cov(f) + E(e) E(f') & \text{because } cov(e,f)=0 \\ &=  \boldsymbol{A\Phi}  \end{align}$ 

<br>

## exercise 3
Use the online material mmm2, tulos 2.4:

$\mu _{f|X} = \mu_f+ \Sigma _{12}  \Sigma _{22}^{-1} (X-\mu_x)$, this is, in our case,

$\begin{align} E(\boldsymbol{f|X} )&= E(f) + cov(f,X) cov(X) ^{-1} (X-E(X)) \\ &=  cov(f,X) cov(X) ^{-1} X & \text{by model, } E(f) = E(X) =0  \\ &=  \boldsymbol{A \Phi} cov(\boldsymbol{X} )^{-1}  \boldsymbol{X}  & \text{the result form exercise 2} \\ &=  \boldsymbol{\Phi}  \boldsymbol{A} ' cov(\boldsymbol{X} )^{-1}  \boldsymbol{X} & \text{becasue } A\Phi=cov(X,f) = (A\Phi)' \end{align}$

<br>

## exercise 5

```{r}
#5a
toimcsvpart=toimcsv[,c("p217","p218","p219","p220","p221",
                       "p222","p223","p224","p225","p226")]
mydata=toimcsvpart

#5b
cor(mydata) #p224 has rather small correlations

#5c
mydata <- mydata[,-8]
library(nFactors)
ev <- eigen(cor(mydata))
ev$values
print("3 eigenvalues greater than 1; Kaiser criterion, 3 factors")

ap <- parallel(subject=nrow(mydata),var=ncol(mydata),
               rep=100,cent=.05)
nS <- nScree(x=ev$values, aparallel=ap$eigen$qevpea)
nS$Components # this also gives Kasiser criterion
print("Cattell's criterion, 2 facotrs")

#5d
fit2 <- factanal(mydata, factors = 2, rotation = "none")
fit3 <- factanal(mydata, factors = 3, rotation = "none")
fit4 <- factanal(mydata, factors = 4, rotation = "none")
fit3
print(paste("The p-values from chi-square, of 2,3 and 4 factors are",
            fit2$PVAL, fit3$PVAL, fit4$PVAL))
print("At least reject 2 factors are sufficient")

#5e
loadings <- fit3$loadings[1:9,]
comm <- NULL
for(i in 1:9){
  comm[i] =loadings[i,1]^2 + loadings[i,2]^2 + loadings[i,3]^2
}
comm
``` 

e. Communalities state how much variance the model explains

f. We can understand loadings just like the regression coefficient, but the factor may be abstract; especially, when not rotated, the factors may not make sense in real life. e.g p222 can be understand as a combined result of 0.504*factor1, 0.419*factor2, -0.73*factor3"

<br>

## exercise 6

```{r}
#6a
fitvari3 <- factanal(mydata, factors=3, rotation="varimax",
                     scores="regression")
#6b 6c
loadings <- fitvari3$loadings[1:9,]
comm <- NULL
for(i in 1:9){
  comm[i] =loadings[i,1]^2 + loadings[i,2]^2 + loadings[i,3]^2
}
comm
loadings
``` 

"Similar with 5f, but this time the factors are independent, and can usually find evidence in real life"

```{r}
#6d
sysScores <- fitvari3$scores
head(sysScores)

invcov <- solve(cov(mydata))
myScores <-  data.matrix(mydata) %*% invcov %*% data.matrix(loadings)
head(myScores) #THIS is non-standardized
print("Obviously, when using the non-standardized, it achieve different result")

barx=colMeans(mydata)
sdx=sqrt(diag(cov(mydata)))
scaledx=sweep(mydata, 2, barx, "-")
standx=sweep(scaledx, 2, sdx, "/")
R1=solve(cov(standx))
myScores <- data.matrix(standx) %*% R1 %*% data.matrix(loadings)
head(myScores)
print("Now it gets the same result as in the R function")

#6e
result <- rbind(colMeans(sysScores),
                c(var(sysScores [,1]), var(sysScores [,2]), var(sysScores [,3])))
row.names(result) <- c("mean", "variance")
result
cor(sysScores)
``` 

<br>

## exercise 7

In Promax-rotated, the factors are not independent, but still similar idea with question 5.


```{r}
fitpromax <- factanal(mydata, factors=3, rotation = "promax", scores="regression")

#7c
sysScores <- fitpromax$scores
result <- rbind(colMeans(sysScores),
                c(var(sysScores [,1]), var(sysScores [,2]), var(sysScores [,3])))
row.names(result) <- c("mean", "variance")
result
cor(sysScores)

#7d
#the following are copied from exercise sheet with minor changes
library(psych)
library(GPArotation)
fitpro3fa<- fa(r = cor(mydata), nfactors= 3,
               rotate = "promax", fm = "ml", scores="regression")
names(fitpro3fa) # objects
fitpro3fa$communalities # compare these to tasks 5 & 6
fitpro3fa$r.scores # correlation matrix of scores
fitpro3fa$Phi # Phi=correlation matrix of factors
fitpro3fa$loadings # A=loadings
fitpro3fa$Structure # Phi*A
# standardization of x variables
barx=colMeans(mydata)
sdx=sqrt(diag(cov(mydata)))
scaledx=sweep(mydata, 2, barx, "-")
standx=sweep(scaledx, 2, sdx, "/")
R1=solve(cov(standx))
# scores with regression method
z=t(fitpro3fa$Structure)%*%R1%*%t(standx)
head(t(z))
# statistics of scores
colMeans(t(z))
cor(t(z))
diag(cov(t(z)))
``` 


