# TILS644 Monimuuttujamenetelmät (Demo 5)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
load("mmm_0.RData")

```

## exercise 1 

$\begin{align} a)\ \bar d _{i\cdot}   &=  \frac{1}{n_i} \sum\limits_{j=1}^{ n_i}   \big( v_1 x _{xy(1)} +v_2 x _{ij(2)}  \big) \\ &= v_1 \cdot \frac{1}{n_i} \sum\limits_{j=1}^{ n_i}   x _{ij(1)} +v_2 \cdot \frac{1}{n_i} \sum\limits_{j=1}^{ n_i}   x _{ij(2)}     \\ &= v_1 \bar x _{i\cdot(1)} + v_2 \bar x _{i\cdot(2)} \\ b)\  \bar d _{\cdot\cdot}  &=  \frac{1}{n} \sum\limits_{i=1}^{ k}  \sum\limits_{j=1}^{ n_i}  d _{ij} \\ &=  \frac{1}{n} \sum\limits_{i=1}^{ k} n_i \bar d _{i\cdot} \\ &=    v_1 \cdot \frac{1}{n} \sum\limits_{i=1}^{ k}  n_i \bar x _{i\cdot(1)}  + v_2 \cdot \frac{1}{n} \sum\limits_{i=1}^{ k}  n_i \bar x _{i\cdot(2)} \\ &=  v_1 \bar x _{\cdot\cdot(1)} + v_2 \bar x _{\cdot\cdot (2)}  \end{align}$

<br>

## exercise 2


$\begin{align} \boldsymbol{v'SS} _{A(x)} \boldsymbol{v'}  &= \begin{bmatrix} v_1 & v_2 \end{bmatrix} \begin{bmatrix}  SS _{A(1)}  & SP _{A(1,2)} \\ SP _{A(1,2)}  & SS _{A(2)}  \end{bmatrix}  \begin{bmatrix} v_1 \\ v_2\end{bmatrix}  \\ &=  v_1^2 SS _{A(1)} +2 v_1 v_2 SP _{A(1,2)}  + v_2^2 SS _{A(2)} \\ &=  \sum\limits_{i} \sum\limits_{j}  \bigg( (v_1 \bar x_{i\cdot(1)} - v_1 \bar x _{\cdot\cdot(1)}  )^2 + 2(v_1 \bar x_{i\cdot(1)} - v_1 \bar x _{\cdot\cdot(1)}  )(v_1 \bar x_{i\cdot(2)} - v_1 \bar x _{\cdot\cdot(2)}  ) + (v_1 \bar x_{i\cdot(2)} - v_1 \bar x _{\cdot\cdot(2)}  )^2 \bigg)  \\ &=  \sum\limits_{i} \sum\limits_{j}  \bigg( (v_1 \bar x_{i\cdot(1)} - v_1 \bar x _{\cdot\cdot(1)}  + v_1 \bar x_{i\cdot(2)} - v_1 \bar x _{\cdot\cdot(2)}   \bigg)^2  \\ &=  \sum\limits_{i}^{ k}  \sum\limits_{j}^{ n_i}   \bigg(  v_1 \bar x _{i\cdot(1)} + v_2 \bar x _{i\cdot (2)} -  v_1 \bar x _{\cdot\cdot(1)} - v_2 \bar x _{\cdot\cdot (2)}    \bigg)^2 \\ &=  SS_A(d)  \end{align}$ 

<br>

## exercise 3

```{r}

# (a)
data1=koulu
data2=data1[,c("V2","V6","V9","V14","V24","V31","V34","V19")]
data2=data2[complete.cases(data2),]
names(data2)=c("gender","luke6","kirj6", "vk16", "mate6", "biol6", "maan6", "hist6")
# (b)
id.boy <- which(data2$gender == 1)
group.mean <- rbind(colMeans(data2[id.boy,]),
                    colMeans(data2[-id.boy,]))
group.mean
``` 

The impressions is that for some subjects (reading, writing, English), different groups have different scores, while no significant differences for other subjects.

<br>

## exercise 4

```{r}
# (a)
data.boy <- data2[id.boy,2:8]; data.girl <- data2[-id.boy,2:8]
cov.boy  <- cov(data.boy); cov.girl <-cov(data.girl)
corr.boy <- cor(data.boy); corr.girl<- cor(data.girl)
cov.boy
cov.girl #correlation matricies are not plotted

library(candisc)
boxM(data2[,-1], data2[,"gender"])
print("We cannot reject that the cov-matricies are similar")

# (b), using R functions 
d.boy <- mahalanobis(data.boy, colMeans(data.boy), cov.boy)
d.girl <- mahalanobis(data.girl, colMeans(data.girl), cov.girl)
# outlier plot
par(mfrow=c(1,2))
plot(d.boy,  ylab="Mahalanobis distance",main="Boy - Outlier")
abline(h = qchisq(0.975,6), col = "red")
plot(d.girl, ylab="Mahalanobis distance",main="Girl - Outlier")
abline(h = qchisq(0.975,6), col = "red")
# qq plot
# qqnormPlot(data.boy)
# qqnormPlot(data.girl)
# qqnormPlot is unsupported in R markdown
``` 

<br>

## exercise 5

```{r}
library(candisc)
data2.mod<-lm(as.matrix(data2[,-1])~data2$gender)
resultcda<-candisc(data2.mod)
resultcda$coeffs.raw
``` 

(a) The first e discriminant function $d_1 = 0.07Luke 6 -1.05 kirj6 -0.42 vk16 + 0.55 mate6-0.38 biol6 + 0.42 maan6 + 0.23 hist6$.
 
Test stat $F=0.735$ with a p-value=.000, which means the function is significant.

```{r}
result <- cbind(resultcda$coeffs.std, resultcda$structure)
colnames(result) <- c("std.coefficients","correlations")
result
group.mean[,3]
``` 

(b) Writing (kirj6) score can discriminate groups, since the std.coef are large, and correlations with the discrimination function is high. We may also see the obvious difference between the mean score among girls and boys.

```{r}
resultcda$means
``` 

<br>

## exercise 6

The R code is below is copied from the question sheet

```{r}
# Division
M<-ncol(data2)
N<-nrow(data2)
train.rows<-sample(1:N,0.5*N)
data2.train<-data2[train.rows,] # training
data2.test<-data2[-train.rows,] # testing
# Formula
features<-data2[,2:M]
form<-as.formula(paste("gender~",paste(names(features),collapse="+")))
# Classification with lda-function and the estimate for the
# classification error
library(MASS)
LDA<-lda(form,data=data2.train)
LDA
LDA.pred<-predict(LDA,newdata=data2.test[,2:M])
Err<-1-sum(LDA.pred$class==data2.test$gender)/nrow(data2.test)
Err
# Classification with qda-function and the estimate for
# the classification error
QDA<-qda(form,data=data2.train)
QDA
QDA.pred<-predict(QDA,newdata=data2.test[,2:M])
Err<-1-sum(QDA.pred$class==data2.test$gender)/nrow(data2.test)
Err
```

The <code>Err</code> value means the probability that the simulated/examed new data belongs to the first group 