# TILS644 Monimuuttujamenetelmät (Demo 6)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
load("mmm_0.RData")
```

## exercise 1

(a) It is data-exploration, trying to clustering (or distinguishing) by certain characteristics (i.e. dimensions).

(b) The key difference is measuring the dissimilarity. The different measures are suitable to different scenarios, since the measures themselves carries application conditions with real-life meanings.

(c) Procrusters analysis is the methodology to find the similarity after scaling, translating, rotation.

<br>

## exercise 2
```{r}
library(mvabund)
data(spider)
Y <- spider$abund
Xenv <- spider$x

matDelta <- matrix(0, nrow = 28, ncol = 28)
for(i in 1:28){
  for(j in 1:28){
    sumsq <- 0
    for(k in 1:12){
      sumsq <- sumsq + (Y[i,k] - Y[j,k])^2
    }
    matDelta[i,j] <- sqrt(sumsq)
  }
}

library(vegan)
vegDelta <- data.matrix(vegdist(Y, method = "euclidean"))
matDelta[1:5,1:5] # only compare a small part of the matrix
vegDelta[1:5,1:5]
``` 

Naturally, they are the same, as the calculation is exactaly the same

<br>

## exercise 3

```{r}
matDelta2 <- matDelta^2 
J <- diag(1,28,28) - (1/28) * matrix(1, 28, 28) 
B <- J %*% matDelta2 %*% J
evB <- eigen(-0.5*B)
Lambda <- diag(evB$values[1:2])
E <- evB$vectors[,1:2]
X <- E %*% sqrt(Lambda)
plot(X)
``` 

<br>

## exercise 4

```{r}
prc <- prcomp(Y)
plot(prc$x)
``` 

Visually, it seems to be 180-degree rotation of the PCoA.

<br>

## exercise 5

```{r}
prc <- procrustes(X, prc$x[,1:2])
prc$ss
plot(prc, kind = 2)
``` 

By the type-2 residual graph and the Procrustes sum of squares (-2.91e-11), it means they are exactly the same (calulation rounding only)

<br>

## exercise 6

```{r}
nmds.eucl <- metaMDS(Y, distance = "euclidean", k = 2, trace = 0)
nmds.bray <- metaMDS(Y, distance = "bray", k = 2, trace = 0)
par(mfrow=c(1,2))
plot(nmds.eucl, type = "t", main = "by Euclidean")
plot(nmds.bray, type = "t", main = "by Bray-curtis")
``` 

Two plots (from different methods) seem similar, but there are obvious differeces in some area. 

The four corners may be catories to 4 clusters. For example, the location 2,4,5,6,7,13,14 have simliar spices (Perdnigr, Auloalbi, Pardpull, Arctlute)

```{r}
par(mfrow=c(1,2))
biplot(nmds.eucl$points, nmds.bray$species)
biplot(nmds.bray$points, nmds.bray$species)
``` 

<br>

## exercise 7

```{r}
ef.eucl <- envfit(nmds.eucl, Xenv, permutations = 999)
ef.bray <- envfit(nmds.bray, Xenv, permutations = 999)
par(mfrow=c(1,2))
plot(nmds.eucl, main = "by Eculidean")
plot(ef.eucl, p.max = 0.1)
plot(nmds.bray, main = "by Bray-curtis")
plot(ef.bray, p.max = 0.1)
``` 

<br>

## exercise 8

```{r}
plot(cca(Y), type = "t")
``` 