---
title: "Demo 2"
author: "Yan Pan | 6 exercises completed | created with R-markdown"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

#exercise 1
###(a) 

$$\begin{align} E(\boldsymbol{Z}) &= E( \boldsymbol{P} ^{T} )  \big( E \boldsymbol{X} - E \boldsymbol{\mu}  \big) \\ &=  E( \boldsymbol{P} ^{T} ) ( \boldsymbol{\mu} - \boldsymbol{\mu} ) \\ &=  \boldsymbol{0}  \\ cov( \boldsymbol{Z} ) &= E  \big( \boldsymbol{P} ^{T}( \boldsymbol{X} -\boldsymbol{\mu} ) \big)  \big( \boldsymbol{P} ^{T}( \boldsymbol{X} -\boldsymbol{\mu} ) \big)^T  \\ &=  E  \bigg(  \boldsymbol{P} ^T (\boldsymbol{X-\mu} ) (\boldsymbol{X-\mu} )^T \boldsymbol{P}  \bigg) \\ &=  \boldsymbol{P} ^T E[(\boldsymbol{X-\mu} ) (\boldsymbol{X-\mu} )^T]  \boldsymbol{P}  \\ &=  \boldsymbol{P} ^T ( \boldsymbol{P\Lambda P}^T ) \boldsymbol{P} \\ &=  \boldsymbol{\Lambda}     \end{align} $$  

###(b)
By the properties of matrix, $tr(\boldsymbol{P\Lambda P}^T ) = tr(\boldsymbol{\Lambda P P} ^T) = tr(\Lambda)$, since $\boldsymbol{P}$ is orthogonal. This proves $tr(\boldsymbol{\Sigma} ) = tr( \boldsymbol{\Lambda} )$.
Further, $tr( \boldsymbol{\Sigma} ) = \sum\limits_{i=1}^{ p}  \lambda_i$  is by property of eigenvalues.

#exercise 2
Calculate the eigenvalues:
 
$\bigg|  \begin{bmatrix} 5-\lambda & 2 \\ 2 & 2- \lambda \end{bmatrix}  \bigg| = \lambda^2 - 7 \lambda + 6 :=0 \ ⇒\  \lambda_1 = 6,\ \lambda_2=1$ and,

the corresponding eigenvectors $\Gamma= \begin{bmatrix} \frac{2}{\sqrt 5} & -\frac{1}{\sqrt 5}   \\ \frac{1}{\sqrt 5} & \frac{2}{\sqrt 5}  \end{bmatrix}$

Therefore the first principal component is $z_1 = \frac{2}{\sqrt 5} x_1 + \frac{1}{\sqrt 5} x_2$, and its variance is $var(Z_1)=\lambda_1=6$  .


#exercise 4
###(a)
```{r}
library(mclust)
data("banknote")
X <- banknote[,2:7]
cov <- cov(X)
Lambda <- diag(eigen(cov)$values)
P <- eigen(cov)$vectors
# set mu matrix
meanValue <- sapply(X, mean)
mu <- t( matrix(rep(meanValue,200), ncol = 200))
# only print first 15 lines
Z <- t(P) %*% t(X-mu)
head(t(Z), n=15)
pairs(t(Z))
``` 

###(b)
Based on the fact that the variances equal eigenvalues

```{r}
barplot(eigen(cov)$values, main = "Variances")
``` 

#exercise 5
```{r}
pka <- prcomp(X)
head(pka$x, n=15)
pairs(pka$x)
screeplot(pka)
``` 

#exercise 6
```{r}
X.new <- X
X.new[,4:6] <- X.new[,4:6] / 100
pka.new <- prcomp(X.new)
vars <- pka.new$sdev^2
firstPK <- vars[1] / sum(vars)
print( paste("The first principal component score explains", firstPK))
``` 

#exercise 7
```{r}
library(MASS)

plotpk <- function(X){
  center <- colMeans(X)
  eigen(cov(X))$vectors
  p1 <- eigen(cov(X))$vectors[,1]
  p2 <- eigen(cov(X))$vectors[,2]
  
  par(pty="s") # force to be square
  plot(X, col = "gray", xlim = c(-10,10), ylim = c(-10,10))
  points(center[1], y=center[2])
  arrows(center[1], center[2], x1 = center[1] + 3*p1[1], y1 = center[2] + 3*p1[2], col = "red")
  arrows(center[1], center[2], x1 = center[1] + 3*p2[1], y1 = center[2] + 3*p2[2], col = "blue")
}

plotpk(mvrnorm(500, c(1,2), matrix(c(4,2,2,9), 2)))
``` 
