# Ryhmittely-, luokittelu- ja regressiomenetelmiä (Demo 6)

```{r echo=F}
load("rlr_0.RData")
```

```{r setup, echo=T, results = 'hide'}
require(tidyverse)
require(MASS)
require(mixtools)
require(reshape2)
require(npmlreg)
```

# 1. Entropy

<blockquote>
With decision tree's entropy, show that $H(Y)$ is maximized when $\pi_l,\ l=1,2,\dots, k$ are equivalent and $H_{max} = \ln (k)$
</blockquote>

Use Langrange function with the restriction $g(\pi_1,\dots,\pi_l) = \sum\limits_{l=1}^k \pi_l -1$, which is from the assumption of decision tree.

$$\begin{aligned} \text{Maximize }\mathcal L(\pi_1,\dots,\pi_k, \lambda) &= - \sum\limits_{l=1}^{k} \pi_l \ln(\pi_l) - \lambda \bigg( \sum\limits_{l=1}^k \pi_l -1 \bigg)
\\ &\Rightarrow 
\begin{cases} \frac{\partial \mathcal L}{\partial \pi_l} &=-(\ln(\pi_l) + 1) - \lambda =0 &\text{for all }l=1,2,\dots, k
\\ \frac{\partial \mathcal L}{\partial \lambda} &= \sum\limits_{l=1}^k \pi_l -1 = 0
\end{cases}
\\ &\Rightarrow 
\begin{cases}\pi_l &= e^{-1-\lambda} &\text{for all }l=1,2,\dots, k
\\ \frac{\partial \mathcal L}{\partial \lambda} &= \sum\limits_{l=1}^k \pi_l -1 = 0
\end{cases}
\\ &\Rightarrow \pi_1 = \pi_2 =\cdots = \pi_l = e^{-1-\lambda} = \frac 1 k
\\ &\Rightarrow H_max = - \sum\limits_{l=1}^k \frac 1 k \ln (\frac 1 k) = - \ln(1/k) = \ln(k)
\end{aligned}$$

Therefore, the entropy is maxized with equal $\pi_l$ and $H_{max} = \ln(k)$

# 2. EM-algorithm

<blockquote>
(Partial proof of EM-algorithm series likelihood is non-descreasing) Let 

$$
R( \boldsymbol\Psi, \boldsymbol\Psi^{(j)}) = \mathbb E \bigg( \log p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi) \ | \ \boldsymbol x, \boldsymbol \Psi^{(j)} \bigg)
$$

Show that $R( \boldsymbol\Psi^{(j+1)}, \boldsymbol\Psi^{(j)}) - R( \boldsymbol\Psi^{(j)}, \boldsymbol\Psi^{(j)}) \leq 0$
</blockquote>

$$\begin{aligned} R( \boldsymbol\Psi^{(j+1)}, \boldsymbol\Psi^{(j)}) - R( \boldsymbol\Psi^{(j)}, \boldsymbol\Psi^{(j)})
&= \mathbb E \bigg( [\log p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi^{(j+1)}) \ | \ \boldsymbol x, \boldsymbol \Psi^{(j)}] -  [\log p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi^{(j)}) \ | \ \boldsymbol x, \boldsymbol \Psi^{(j)}] \bigg)
\\ &= \mathbb E \bigg( \log  \frac{p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi^{(j+1)}) \ | \ \boldsymbol x, \boldsymbol \Psi^{(j)}}{ p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi^{(j)}) \ | \ \boldsymbol x, \boldsymbol \Psi^{(j)}} \bigg)
\\ &\leq \log \bigg(\mathbb E  \frac{p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi^{(j+1)}) \ | \ \boldsymbol x, \boldsymbol \Psi^{(j)}}{ p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi^{(j)}) \ | \ \boldsymbol x, \boldsymbol \Psi^{(j)}} \bigg)
\\ &\leq \log \bigg(\mathbb E  \frac{p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi^{(j+1)}) \ | \ \boldsymbol x, \boldsymbol \Psi^{(j)}}{ p( \boldsymbol Z| \boldsymbol x, \boldsymbol \Psi^{(j)})}  \bigg)
\\ &= \log 1= 0
\end{aligned}$$

# 3. Mixed Distribution Plot

```{r rlr603a}
par(mfrow=c(1,2))
hist(hoikkarapumjat$Height, breaks = 10)
hist(hoikkaeirapumjat$Height, breaks = 10)

normal2koe <-alldist(Height~1,data=hoikkarapumjat, k=2, plot.opt = 0)
summary(normal2koe)

mixed_distribution <- function(x){
  em_means <- normal2koe$mass.points
  em_sd <- normal2koe$sdev$sdev
  em_pis <- normal2koe$masses
  d1 <- dnorm(x, mean = em_means[1], sd = em_sd)
  d2 <- dnorm(x, mean = em_means[2], sd = em_sd)
  y <- em_pis[1] * d1 + em_pis[2] * d2
  return(list(p1 = d1, p2 = d2, y = y))
}

x = 100:800
y = numeric(701)
for (i in 1:701) y[i] = mixed_distribution(x[i])$y
plot_data = data.frame(x=x, y=y)

ggplot() + 
  geom_histogram(data=hoikkarapumjat, aes(x = Height, y = ..density..), bins = 10) +
  geom_line(data=plot_data, aes(x, y))
```

# 4. Clustering

```{r rlr604a}
koe_clusters <- ifelse(normal2koe$post.prob[,1] > normal2koe$post.prob[,2], 1, 2 )
koe_kmeans   <- kmeans(hoikkarapumjat$Height, 2)$cluster
table(koe_clusters, koe_kmeans)
```

It seems to have exactly same results!

# 5. Number of Mixture Components

```{r rlr605a}
thefun <- function(kk) {
  mixed <- alldist(Width~1,data=hoikkarapumjat, k=kk, 
               verbose = F, plot.opt = 0)
  return(list(mixed$disparity, mixed$mass.points))
}

# use width
lapply(1:6, thefun)
```

Though $k=6$ gives smallest disparity, 2 mixtures seem to be more practical as the centers are very closed to each other in more mixture cases.

# 7. Multi-dimensional Mixed Distribution 

```{r rlr607a}
names(hoikkaeirapumjat) <- names(hoikkarapumjat)
all_data <- rbind(hoikkaeirapumjat, hoikkarapumjat)
melt(all_data) -> a

all_data$ryhma <- "kontrolli"
all_data$ryhma[1:181] <- "koe"

ggplot(reshape2::melt(all_data, id.vars = "ryhma"))+ 
  geom_boxplot(aes(y = value, color = ryhma)) +
  facet_wrap(~variable, scales = "free")

init_mu <- list( c(200, 200, 1e4, 1000, 200, 200, 200),
                 c(650, 650, 2e4, 2000, 600, 750, 750),
                 c(300, 300, 3e4, 1200, 300, 300, 300))
# tulosrapu  <- mvnormalmixEM(hoikkarapumjat  , mu = init_mu, k=3)
# tuloseirapu<- mvnormalmixEM(hoikkaeirapumjat, k=3)
```

# 8. Find Best Solution

```{r rlr608a}
# code referenced (J. Isoketo & J. Itkonen, 2012)
tulos.EM = function(data) {
testiLoglik = data.frame(loglik= -1000000000)
loglik = -1000000
tulos.paras = NULL
n = 1
while (n <= 30) {
tulos = tryCatch(mvnormalmixEM(data,k=3), error=function(e) testiLoglik)
#sovitus ei aina onnistu
if (tulos$loglik > loglik) {
loglik = tulos$loglik
tulos.paras = tulos }
n = n+1
}
tulos.paras
}

# end ref

# tulos.rapu.paras = tulos.EM(data=hoikkarapumjat)

```

# 9. N<<p: Uncorrelated Variables

```{r rlr609a}
n <- sum(cor(training[,-1]) > 0.75) - ncol(training)
cat("There are",n, "pairs of highly correlated vars")
lda_fit <- lda(class ~ ., data=training)
lda_pred <- predict(lda_fit, newdata = testing[,-1])
table(lda_pred$class, testing$class)
```

# 10. N<<p: PCA

```{r rlr610a}
training_pca <- prcomp(training[,-1])$x[, 1:20]
testing_pca  <- prcomp(testing [,-1])$x[, 1:20]

training_pca <- as.data.frame(training_pca)
training_pca$class <- training$class
lda_pca <- lda(class ~., data=training_pca)
pred_pca <- predict(lda_pca, newdata = as.data.frame(testing_pca))
table(pred_pca$class, testing$class)

```