---
author: "PAN Yan"
---

```{r setup, echo=F, result = 'hide'}
require(magrittr)
require(cluster)
require(fpc)
require(clue)
```

Working with Bugs-data (Pohjaeläinaineisto). There are a few modifications from training R code.

# 1. Reading data

```{r rlr301a}
bugs<-read.table("bugs.dat",header=TRUE)
bugs %>% str
```

# 2. Filter data

```{r rlr302a}
index1<-which( bugs$Species=="Baetisrhodani")
index2<-which( bugs$Species=="Hydropsychepellucidulla")
osadata<-bugs[c(index1,index2),]
osadata$Species<-factor(osadata$Species, labels=c("Baetis","Hydro"))
osadata<-osadata[, c(1,2,8)]
names(osadata)
```

# 3. Training and Testing Datasets

```{r rlr303a}
# randomly divided into two subsets
M<-ncol(osadata)
N<-nrow(osadata)
set.seed(123)
train.rows<-sample(1:N, 0.5*N)
osa.train<-osadata[train.rows,]
osa.test<-osadata[-train.rows,]
```

# 4. Check Data

```{r rlr304a}
  # a lot of modifications from Harj3
baetis<-osa.train[osa.train$Species=="Baetis",]
summary(baetis)
hydro<-osa.train[osa.train$Species=="Hydro",]
summary(hydro)

require(ggplot2); require(reshape2)
osa.train %>% 
  melt("Species") %>% # long format for easy plotting
  ggplot() + geom_boxplot(aes(y = value, color = Species)) + 
  facet_wrap(~variable, scales = "free_y")

```

# 5. Classification of Test Data

```{r rlr305x}
  # prepare data structure
osa.test$SpeciesInt <- as.numeric(osa.test$Species)
osa.test$SpeciesLk <- rep(0, nrow(osa.test))
  # key value from training
training1 <- osa.train[osa.test$SpeciesInt == 1, 2:3]
training2 <- osa.train[osa.test$SpeciesInt == 2, 2:3]
m1 <- colMeans(training1)
m2 <- colMeans(training2)
cov1 <- cov(training1)
cov2 <- cov(training2)
```

(a) Euclidean Distance

```{r rlr305a}
osa.test.euclidean <- osa.test
for(i in 1:nrow(osa.test.euclidean)){
  d1 <- dist(rbind(osa.test.euclidean[i, 2:3], m1), method = "euclidean")
  d2 <- dist(rbind(osa.test.euclidean[i, 2:3], m2), method = "euclidean")
  if(d1 <= d2) osa.test.euclidean$SpeciesLk[i] <- 1
  else osa.test.euclidean$SpeciesLk[i] <- 2
}
table(osa.test.euclidean$SpeciesInt, osa.test.euclidean$SpeciesLk)
```

(b) Mahalanobis Distance

```{r rlr305b}
osa.test.mahalanobis  <- osa.test
for(i in 1:nrow(osa.test.mahalanobis )){
  d1 <- mahalanobis(osa.test.mahalanobis [i, 2:3], m1, cov1)
  d2 <- mahalanobis(osa.test.mahalanobis [i, 2:3], m2, cov2)
  if(d1 <= d2) osa.test.mahalanobis $SpeciesLk[i] <- 1
  else osa.test.mahalanobis$SpeciesLk[i] <- 2
}
table(osa.test.mahalanobis$SpeciesInt, osa.test.mahalanobis$SpeciesLk)
```

(c) 2-dimensional Normal Density with Same Prior

```{r rlr305c}
require(mvtnorm)

osa.test.mvnorm  <- osa.test
for(i in 1:nrow(osa.test.mvnorm )){
  p1 <- dmvnorm(osa.test.mvnorm [i, 2:3], m1, cov1)
  p2 <- dmvnorm(osa.test.mvnorm [i, 2:3], m2, cov2)
    # greater p implies classification
  if(p1 <= p2) osa.test.mvnorm $SpeciesLk[i] <- 2
  else osa.test.mvnorm$SpeciesLk[i] <- 1
}
table(osa.test.mvnorm$SpeciesInt, osa.test.mvnorm$SpeciesLk)
```

(d) 2-dimensional Normal Density with Training Probability as Prior

```{r rlr305d}
prior1 <- sum(osa.train$Species == "Baetis")/nrow(osa.train)
prior2 <- sum(osa.train$Species == "Hydro")/nrow(osa.train)

  # use posterior directly, instead of derived conditions
osa.test.mvnormprior <- osa.test
for(i in 1:nrow(osa.test.mvnormprior)){
  p1 <- dmvnorm(osa.test.mvnormprior[i, 2:3], m1, cov1)
  p2 <- dmvnorm(osa.test.mvnormprior[i, 2:3], m2, cov2)
  post1 <- prior1 * p1 / (prior1 * p1 + prior2 * p2)
  post2 <- prior2 * p2 / (prior1 * p1 + prior2 * p2)

    # greater p implies classification
  if(post1 <= post2) osa.test.mvnormprior$SpeciesLk[i] <- 2
  else osa.test.mvnormprior$SpeciesLk[i] <- 1
}
table(osa.test.mvnormprior$SpeciesInt, osa.test.mvnormprior$SpeciesLk)

```

(e) 2-dimensional Normal Density by Naive Bayesian

```{r rlr305e}
osa.test.naivebayes <- osa.test
cov1.diag <- matrix(c(cov1[1,1], 0, 0, cov1[2,2]), ncol = 2)
cov2.diag <- matrix(c(cov2[1,1], 0, 0, cov2[2,2]), ncol = 2)

for(i in 1:nrow(osa.test.naivebayes)){
  p1 <- dmvnorm(osa.test.naivebayes[1, 2:3], m1, cov1.diag)
  p2 <- dmvnorm(osa.test.naivebayes[i, 2:3], m2, cov2.diag)
  post1 <- prior1 * p1 / (prior1 * p1 + prior2 * p2)
  post2 <- prior2 * p2 / (prior1 * p1 + prior2 * p2)

    # greater p implies classification
  if(post1 <= post2) osa.test.naivebayes$SpeciesLk[i] <- 2
  else osa.test.naivebayes$SpeciesLk[i] <- 1
}
table(osa.test.naivebayes$SpeciesInt, osa.test.naivebayes$SpeciesLk)

```

# 6. LDA function

```{r rlr306a}
require(MASS)
  # seems to be corresponding to Euclidean
LDA<-lda(Species ~.,data=osa.train, prior=c(0.50,0.50))
LDA.pred<-predict(LDA,newdata=osa.test)
table(osa.test$Species,LDA.pred$class)
```

# 7. QDA function

```{r rlr307a}
QDA<-qda(Species ~.,data=osa.train, prior=c(0.50,0.50))
QDA.pred<-predict(QDA,newdata=osa.test)
table(osa.test$Species, QDA.pred$class)
```

# 8. Consider Loss Function using QDA result

```{r rlr308a}
post1.qda <- QDA.pred$posterior[,1]
post2.qda <- QDA.pred$posterior[,2]
loss1on2 <- 1
loss2on1 <- 2

osa.test.qdaloss <- osa.test

for(i in 1:nrow(osa.test.qdaloss)){
  risk1 <- loss1on2 * post2.qda[i]
  risk2 <- loss2on1 * post1.qda[i]

    # greater p implies classification
  if(risk1 <= risk2) osa.test.qdaloss$SpeciesLk[i] <- 1
  else osa.test.qdaloss$SpeciesLk[i] <- 2
}
table(osa.test.qdaloss$SpeciesInt, osa.test.qdaloss$SpeciesLk)


```