---
title: ""
author: "PAN Yan"
output:
  html_document:
    df_print: paged
    theme: readable
---

<style>
h1,h2,h3{
  font-family: Georgia,"Times New Roman",Times,serif !important;
  padding-top: 20px;
}
blockquote{
  font-size: 16px; !important;
}
</style>

```{r echo=F, results='hide'}
require(magrittr)
require(cluster)
require(fpc)
```

(R exercises attached below the theoretic ones)

## 8. Theory: weight and standardization

<blockquote>
For $d_j(x_{ij},\ x_{i'j}) = (x_{ij} - x_{i'j})^2$, show that

$$
\bar d_j = \frac 1 {N^2} \sum\limits_{i=1}^N \sum\limits_{i'=1}^N (x_{ij} - x_{i'j})^2 =  
2 \frac 1 {N} \sum\limits_{i=1}^N  (x_{ij} - \bar x_j)^2 = 2s_j^2
$$
</blockquote>

$$\begin{aligned} \bar d_j
&= \frac 1 {N^2} \sum\limits_{i=1}^N \sum\limits_{i'=1}^N (x_{ij} - \bar x_j + \bar x_j - x_{i'j})^2
\\ &= \frac 1 {N^2} \bigg(
N\sum\limits_{i=1}^N (x_{ij} - \bar x_j)^2 +
N\sum\limits_{i'=1}^N (x_{i;j} - \bar x_j)^2 +
2 \sum\limits_{i=1}^N \sum\limits_{i'=1}^N (x_{ij} - \bar x_j) (x_{i'j} - \bar x_j)
\bigg)
\\ &= \frac 2 {N^2} \bigg(
N\sum\limits_{i=1}^N (x_{ij} - \bar x_j)^2 +
\sum\limits_{i=1}^N \sum\limits_{i'=1}^N (x_{ij} - \bar x_j) (x_{i'j} - \bar x_j) \bigg)
\\ &= \frac 2 {N^2} 
\sum\limits_{i=1}^N  \bigg( N(x_{ij} - \bar x_j)^2 +
\sum\limits_{i'=1}^N (x_{ij} - \bar x_j) (x_{i'j} - \bar x_j) \bigg)
\\ &= \frac 2 {N^2} 
\sum\limits_{i=1}^N  \bigg( N(x_{ij} - \bar x_j)^2 + (x_{ij} - \bar x_j)
\underbrace{ \sum\limits_{i'=1}^N  (x_{i'j} - \bar x_j) } _{=0} \bigg)
\\ &= 2 \frac 1 {N} \sum\limits_{i=1}^N  (x_{ij} - \bar x_j)^2 = 2s_j^2
\end{aligned}$$

(question b not done)

## 1. Get Familiar with Data

<blockquote>
a) How is altruism defined in the thesis?
</blockquote>

They use the first principle component of (transformed) survey result.

<blockquote>
b) read data <code>altruism.dat</code> and correct B18E data.
</blockquote>

```{r rlr101a}
options(stringsAsFactors = T)
aineisto <- read.table("altruismi.dat")
str(aineisto)
aineisto$B18E[aineisto$B18E == 133.9] <- 13.9
```

## 2. Manipulate Data

<blockquote>
Processing of nominal variable 
</blockquote>

All nominal variables are ordinal, and therefore, we need to specify the ordinal of these factor-variable.

```{r rlr102a}
aineisto$B8 <- factor(aineisto$B8, ordered = T,
                      levels = c("No", "Yes"))
aineisto$B12 <- factor(aineisto$B12, ordered = T,
                       levels = c("never", "sometimes", "regularly"))
aineisto$B13 <- factor(aineisto$B13, ordered = T,
                       levels = c("no", "yes"))
aineisto$B14 <- factor(aineisto$B14, ordered = T,
                       levels = c("I don't like", "never", "sometimes" ,"regularly"))
aineisto$B16 <- factor(aineisto$B16, ordered = T,
                       levels = c("I don't like", "never", "sometimes" ,"regularly"))


diss <- daisy(aineisto)
summary(diss)
```

## 3. pam- and cluster.varstats functions

```{r rlr103a}
# parse a dissimilarity object will automatically use x as diss
res_pam <- pam(x = diss, # the input matrix, observation or diss
               k = 5, # specified number of clusters
               diss = TRUE) # whether input is diss or not
# only show partil resutls
res_pam$medoids
```

```{r rlr103b, eval=F}
require(fpc)
help("cluster.varstats")
```

## 4. Clustering altruism data using K-medoids of Gower-distance

```{r rlr104a}
# complete case
aineisto_complete <- na.omit(aineisto)
# daisy gower distance
diss_gower <- daisy(aineisto_complete, metric = "gower")
# K-medoids
cls_3 <- pam(diss_gower, 3, diss = T)
cls_3$medoids
```

```{r rlr104b, eval = F}
# report using cluster.varstats
cluster.varstats(clustering = cls_3$clustering,
                 vardata = aineisto_complete,
                 tablevar = c(2, 4, 5, 6, 8),
                 catvar = c(1,3,7, 9:17))
```

result is hidden due to excessive lenght.

## 5. Clustering using K-medoid with Gower-distance and weight

```{r rlr105a}
# daisy gower distance
diss_gower_weighted <- daisy(aineisto_complete, 
                             metric = "gower",
                             weights = runif(17))
# K-medoids
cls_3_weighted <- pam(diss_gower_weighted, 3, diss = T)
cls_3_weighted$medoids
```

The results are surely different if weights are assigned significantly.

## 6. K-means Function

```{r rlr106a}
# kmeans does not accept factors, manually fix it
aineisto_c <- aineisto_complete
lapply(list(2, 4, 5, 6, 8), function(x)
  {aineisto_c[,x] <- as.numeric(aineisto_complete[,x])})

# now call kmeans
kmeans(aineisto_c[,1], centers = 3)
```

## 7. K-means clustering of generic data

```{r rlr107a, eval=F}
nci <- read.table(url("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/nci.data.csv"), sep=",",row.names=1, header=TRUE)

label <- scan(url("https://web.stanford.edu/~hastie/ElemStatLearn/datasets/nci.label.txt"), what="char")

save(nci, label, file = "nci.rdata")
```

```{r rlr107b}
  # load cached data (improve r-markdown performance)
load("nci.rdata")
  # transpose
nci_df <- data.frame(t(nci))
  # k-means
nci_k_means <- kmeans(nci_df, 3)
  # cross tab
table(label, nci_k_means$cluster)
```

