# Ryhmittely-, luokittelu- ja regressiomenetelmiä (Demo 2)

```{r echo=F}
load("rlr_0.RData")
```

```{r setup, echo=T, results='hide'}
require(magrittr)
require(cluster)
require(fpc)
require(clue)
```

# 1. Understand R-functions
<blockquote>
Get familiar with functions `agnes` `plot.agnes` and `summary.agnes`
</blockquote>
`agnes` computes agglomeration hierarchical clustering of the data-set.

```{r rlr201a}
require(cluster)
# quoted from help page
data(votes.repub) # sample data
agn1 <- agnes(votes.repub, 
              metric = "manhattan",  # can be "euculidean"
              stand = TRUE)          # standardize before?

plot(agn1, which.plots = 2) # this will call plot.agnes

summary(agn1) # this will call summary.agnes

```



# 2. Understand R-functions

<blockquote>
Get familiar with functions `dist` `diana`, `plot.diana` and `cluster.stats`.
</blockquote>

`diana` computes a divisive hierarchical clustering of the data-set returning an object of class diana.

The result and plot is in the same form as `agnes`


```{r rlr202a}
# dist(votes.repub, method = "euclidean", p = 2) # other variables are not important 
# we can pass the result of dist instead of data to diana
d  <- dist(votes.repub, method = "manhattan")
dv <- diana(d, stand = TRUE)
plot(dv, which.plots = 2)
cluster <- dv$order
require(fpc)
cluster.stats(d, cutree(dv,2)) %>% unlist()
```



# 3. Divisive methods and dendrograms

<blockquote>
Using the matrix $X$ given in lecture notes and choose proper dendrogram for distance matrix.
</blockquote>
```{r rlr203a}
x <- matrix(c(1,1,2,1,5,4,6,5,6.5,6), byrow=T, ncol=2)
dist(x, method = "manhattan") %>%
  diana() %>%
  plot(which.plots = 2)
# same as agnes(x, metric = "manhattan")

```


# 4. Altruism data set clustering by hierarchical methodology

<blockquote>
Using different linkage function to perform clustering with manhattan distance.
</blockquote>

```{r rlr204a}
altruismi <- altruismi_corrected
# perform clustering | results are in a list
clusters <- lapply(c("average", "single", "complete", "ward"),
                  function(m) agnes(altruismi, metric="manhattan", method=m, keep.diss = T))
clusters[[5]] <-diana(altruismi, metric = "manhattan", keep.diss = T)
names(clusters) <- c("average", "single", "complete", "ward", "diana")

# plots
par(mfrow=c(2,3))
lapply(clusters, plot, which.plots = 2, main = "")
```

Based on the graph, it seems that the agglomerative nesting by complete linkage and ward, together with divisive methods are relatviely more proper.

```{r rlr204b}
require(clue)
lapply(clusters, function(cl) cl_dissimilarity(cl, cl$diss,  "cophenetic"))
```

Therefore, agglomerative nesting with Ward method is the best in the sense of Cophenetic correlation.


# 5. Random test of Ward method

Based on random position hypothesis with DISCRETE uniform distribution.

</blockquote>
```{r rlr205a}
simAltruismi <- function(originalData = altruismi){
  simData <- originalData
  N <- nrow(originalData)
  for(col_i in 1:ncol(originalData)){
    simData[ ,col_i] <- sample(originalData[, col_i], N, replace = T)
  }
  simData
}

testWard <- function(m = 100){
  cpcc <- numeric(m)
  for (i in 1:m) {
    simData <- simAltruismi()
    result  <- agnes(simData, metric="manhattan", method="ward", keep.diss = T)
    cpcc[i] <- cl_dissimilarity(result, result$diss, "cophenetic")
  }
  cpcc
}

hist(testWard())
```


# 7. Choice of $k$
<blockquote>
Select the most suitable $k$ in case of Ward method.
</blockquote>

```{r rlr207a}
agnes_ward <- clusters[[4]] # from previous exercises
# note list starts from 2 to 8
cl_ward_gamma <- lapply(2:8, function(k) 
  cluster.stats(agnes_ward$diss, cutree(agnes_ward, k))$pearsongamma)
cl_ward_gamma %>% unlist
```
So, choose $k=3$

<blockquote>
Select the most suitable $k$ in case of K-medoids method.
</blockquote>
```{r rlr207b}
diss <- dist(altruismi, method = "manhattan")
cl_pam_gamma <- lapply(2:8, function(k)
  cluster.stats(diss, pam(altruismi, k)$clustering)$pearsongamma)
cl_pam_gamma %>% unlist
```

In this case, $k = 6$ is the best.

<blockquote>
Which one of the above is better?
</blockquote>

Ward-method seems to be nicer due to smaller normalized Huber Gamma (Pearson Gamma).


# 8. Choice of the method, if fix the group number to be 3

Use Huber Gamma to measure

```{r rlr208a}
# hierarchical results from exercise 4


gammas <- lapply(clusters, function(cl)
  cluster.stats(diss, cutree(cl, 3))$pearsongamma)
gammas[[6]] <- cl_pam_gamma[4] # from the previous
names(gammas) <- c("average", "single", "complete", "ward", "diana", "pam")

gammas
```

The best is still Ward method


# 9. Gene data set

<blockquote>
Using hierarchical methods and Euclidean distance.
</blockquote>
```{r rlr209as}
load("nci.rdata")
nci_df <- data.frame(t(nci))
# Euclidean is the default value for metric
clusters <- lapply(c("average", "single", "complete", "ward"),
                  function(m) agnes(nci_df, method=m, keep.diss = T))
clusters[[5]] <-diana(altruismi, keep.diss = T)
names(clusters) <- c("average", "single", "complete", "ward", "diana")

par(mfrow = c(2,3))
lapply(clusters, plot, which.plots = 2, main = "")

```
