# Ryhmittely-, luokittelu- ja regressiomenetelmiä (Demo 4)

```{r echo=F}
load("rlr_0.RData")
```

```{r setup, echo=T, results = 'hide'}
require(magrittr)
require(cluster)
require(fpc)
require(clue)
require(tidyverse)
```

Working with Bugs-data (Pohjaeläinaineisto). 

Unlike the training code, this exercise using TidyVerse style analysis flow.

# 1. Reading data and modifications

```{r rlr401a}
# filtering and subsetting using tidyverse
osadata <- bugs %>%     
  filter(Species %in% c("Baetisrhodani", "Hydropsychepellucidulla")) %>%
  dplyr::select("Species", "Area", "Height")
osadata

# sample train and test data obs
set.seed(123)
train_rows <- sample(1:nrow(osadata), nrow(osadata)/2)
osa.train <- osadata[ train_rows, ]
osa.test  <- osadata[-train_rows, ]

# plotting using ggplot and long data
osadata$trainORtest <- "TEST"
osadata$trainORtest[train_rows] <- "TRAIN"
osadata %>%
  reshape2::melt(c("Species", "trainORtest")) %>%
  ggplot() +
  geom_boxplot(aes(y = value, color = Species)) +
  facet_wrap(variable ~ trainORtest, scales = "free_y") +
  theme(legend.position = "bottom", aspect.ratio = 1)

```

# 2. Plot Decision Areas

(Prepare data structure)

```{r rlr402a}
  # prepare data structure
osa.test$SpeciesLk <- rep(0, nrow(osa.test))
  # key value from training
training1 <- osa.train[osa.test$Species == "Baetisrhodani", 2:3]
training2 <- osa.train[osa.test$Species == "Hydropsychepellucidulla", 2:3]
m1 <- colMeans(training1)
m2 <- colMeans(training2)
cov1 <- cov(training1)
cov2 <- cov(training2)

osadata %>% ggplot(aes(Area, Height)) + geom_point()
```

## (1) by Euclidean distance

```{r rlr402b}
osa.test.euclidean <- osa.test
for(i in 1:nrow(osa.test.euclidean)){
  d1 <- dist(rbind(osa.test.euclidean[i, 2:3], m1), method = "euclidean")
  d2 <- dist(rbind(osa.test.euclidean[i, 2:3], m2), method = "euclidean")
  if(d1 <= d2) osa.test.euclidean$SpeciesLk[i] <- 1
  else osa.test.euclidean$SpeciesLk[i] <- 2
}
osa.test.euclidean %>% 
  ggplot(aes(Area, Height, color = as.factor(SpeciesLk))) +
  geom_point() + ggtitle("Euclidean")
```


(b) Mahalanobis Distance with polarized covariance

```{r rlr402c}
osa.test.mahalanobis  <- osa.test
covp<-((nrow(training1)-1)*cov1+(nrow(training2)-1)*cov2)/(nrow(osa.train)-2)

for(i in 1:nrow(osa.test.mahalanobis )){
  d1 <- mahalanobis(osa.test.mahalanobis [i, 2:3], m1, covp)
  d2 <- mahalanobis(osa.test.mahalanobis [i, 2:3], m2, covp)
  if(d1 <= d2) osa.test.mahalanobis $SpeciesLk[i] <- 1
  else osa.test.mahalanobis$SpeciesLk[i] <- 2
}

osa.test.mahalanobis %>% 
  ggplot(aes(Area, Height, color = as.factor(SpeciesLk))) +
  geom_point() + ggtitle("Mahalanobis Dsitance and COV_P")
```

(c) Mahalanobis Distance with Class-wise Covaraince

```{r rlr403d}
osa.test.m2  <- osa.test

for(i in 1:nrow(osa.test.m2)){
  d1 <- mahalanobis(osa.test.m2 [i, 2:3], m1, cov1)
  d2 <- mahalanobis(osa.test.m2 [i, 2:3], m2, cov2)
  if(d1 <= d2) osa.test.m2$SpeciesLk[i] <- 1
  else osa.test.m2$SpeciesLk[i] <- 2
}

osa.test.m2 %>% 
  ggplot(aes(Area, Height, color = as.factor(SpeciesLk))) +
  geom_point() + ggtitle("Mahalanobis Dsitance and Class-wise Covariance")
```