---
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
    df_print: kable
    theme: 'readable'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(geoR)
options(geoR.messages = FALSE)
```

# 2-2 Kriging without and with nugget effect

> without nugget

```{r d22a}
esim <- data.frame(x = c(0.2, 0.5, 0.4, 0.9),
                   y = c(0.2, 0.1, 0.5, 0.8),
                   dat = c(12, 6, 10,2 ))
y.geo<-as.geodata(matrix(c(esim$x,esim$y,esim$dat),ncol=3))
simplek <- krige.control(type.krige="sk",beta=5.0, 
                       cov.model="exponential",cov.pars=c(1,0.2))
# Ennustus uuteen paikkaan (0.5,0.4):
tul<-krige.conv(y.geo,locations=c(0.5,0.4),krige=simplek)
cat("predict & var: ", tul$predict, tul$krige.var)

#using manual calculation
cfunc <- function(r) exp(-5*r)
s <- matrix(c(0.2,0.5,0.4,0.9,0.2,0.1,0.5,0.8),ncol=2)
y <- matrix(c(12,6,10,2),ncol=1)
u <- y-5.0
C <- cfunc(as.matrix(dist(s,diag=TRUE)))

s0 <- matrix(c(0.5,0.4),ncol=1)
s.aug <- matrix(c(s[,1],s0[1,1],s[,2],s0[2,1]),ncol=2)
cs <- as.matrix(cfunc(as.matrix(dist(s.aug))[5,1:4]))
lambda0 <- solve(C)%*%cs
hat.u <- t(lambda0)%*%u
hat.y <- hat.u+5.0
k.var <- 1-t(cs)%*%solve(C)%*%cs
cat("predict & var:", hat.y, k.var)
```

> with nuggest = 0.8

```{r d22b}
simpleknugget<-krige.control(type.krige="sk",beta=5.0,nugget=0.8, 
                             cov.model="exponential",cov.pars=c(1,0.2))
tulnugget<-krige.conv(y.geo,locations=c(0.5,0.4),krige=simpleknugget)
cat(tulnugget$predict, tulnugget$krige.var)

#manually
cfunc <- function(r)  exp(-5*r)
s<-matrix(c(0.2,0.5,0.4,0.9,0.2,0.1,0.5,0.8),ncol=2)
y<-matrix(c(12,6,10,2),ncol=1)
u<-y-5.0
C<-cfunc(as.matrix(dist(s,diag=TRUE)))
# lisataan nugget diagonaalille
yksikko= diag(nrow = 4)
nugget=0.8
C=C +nugget*yksikko

s0<-matrix(c(0.5,0.4),ncol=1)
s.aug<-matrix(c(s[,1],s0[1,1],s[,2],s0[2,1]),ncol=2) # jarjestys vaihdettu s0:ssa
cs<-as.matrix(cfunc(as.matrix(dist(s.aug))[5,1:4]))
lambda0<-solve(C)%*%cs
hat.u<-t(lambda0)%*%u
hat.y<-hat.u+5.0
k.var<-1+nugget-t(cs)%*%solve(C)%*%cs # tausq+sigmasq
cat(hat.y, k.var)
```

> more general methods

```{r d22c}
# Tavanomainen kriging valmiilla funktiolla, kun nugget=0.8 mukana:  
oknugget <- krige.control(type.krige="ok",nugget=0.8, 
                          cov.model="exponential",cov.pars=c(1,0.2))
tul<-krige.conv(y.geo,locations=c(0.5,0.4),krige=oknugget)
cat(tul$predict, tul$krige.var)

# universaali kriging valmiilla funktiolla, kun nugget=0.8 mukana:  
uknugget<-krige.control(type.krige="ok",nugget=0.8, 
                        cov.model="exponential",cov.pars=c(1,0.2),
                        trend.d="1st", trend.l="1st")
tul<-krige.conv(y.geo,locations=c(0.5,0.4),krige=uknugget)
cat(tul$predict, tul$krige.var)
```

# 2-3 Variogram Fitting

On kriging the calcium material data (`ca20`). Check the first and second order trend, using a) `variofit` function and b) `likfit` function, and compare the result

> Variogram Cloud and anisotorpy

```{r d23a}
par(mfrow = c(1,2))
plot(variog(ca20, option='cloud'),
     pch='.',cex=0.5, main = "cloud variogram")
a <- variog(ca20, option='smooth', band=100)
lines(a$u,a$v,col=2)
plot(variog(ca20), main = "ordinary")

# anisotropy
plot(variog4(ca20)) # or use variog to calculate each direction
plot(variog4(ca20, trend = '1st'))
plot(variog4(ca20, trend = "2nd"))

# more detailed semivariance plot
mySemivario <- function(trend = "cte"){
  a1<-variog(ca20, option='smooth', direction=0,band=100,trend=trend)
  a2<-variog(ca20, option='smooth', direction=pi/4,band=100,trend=trend)
  a3<-variog(ca20, option='smooth', direction=pi/2,band=100,trend=trend)
  a4<-variog(ca20, option='smooth', direction=3*pi/4,band=100,trend=trend)
  n.o<-c("0","45","90","135")
  ymax <- max(c(a1$v, a2$v, a3$v, a4$v), na.rm = TRUE)
  plot(a1,type='l',xlab='distance',ylab='semivariance',ylim = c(0, ymax),
       main = "smoothed variograms of residuals")
  lines(a2$u,a2$v,lty=2)
  lines(a3$u,a3$v,lty=3,lwd=2)
  lines(a4$u,a4$v,lty=4)
  legend(0, ymax, legend = n.o,lty = c(1:4),lwd=c(1,1,2,1))
}

mySemivario("1st") #eg
```

> Variofit

```{r d23b}
# for example 1st trend
ca20.bin1st <- variog(ca20, option="bin",
                      estimator.type="classical",
                      trend='1st', max.dist = 500)
plot(ca20.bin1st)

# Parametrisen mallin sovitus variofitin kautta
varioe1st <- variofit(ca20.bin1st, ini.cov.pars =c(80,20), 
                      cov.model="exponential",
                       fix.nugget=FALSE,nugget=40)
varioe1st
variog1st <- variofit(ca20.bin1st, ini.cov.pars =c(80,20), 
                      cov.model="gaussian",
                      fix.nugget=FALSE,nugget=40)
variog1st
varios <- variofit(ca20.bin1st, ini.cov.pars =c(80,20), 
                   cov.model="spherical",
                   fix.nugget=FALSE,nugget=40)
varios 

plot(ca20.bin1st)
lines(variog1st, col ="red")
lines(varioe1st, col="blue")
lines(varios, col="green")

# constuct data for prediction
xlim<-range(ca20$borders[,1]); 
ylim<-range(ca20$borders[,2])
xgrid<-seq(xlim[1],xlim[2],l=100)
ygrid<-seq(ylim[1],ylim[2],l=100)
pred.grid <- expand.grid(xgrid,ygrid)
# universal kriging
krige.par <- krige.control(type.krige='ok', 
                           cov.pars=varioe1st$cov.pars, 
                           nugget=varioe1st$nugget,
                           cov.model="exp",trend.d="1st", trend.l="1st")
k<-krige.conv(ca20,locations=pred.grid,krige=krige.par)


image(k)
points(ca20, pch=20,cex.max=1,pt.divide='equal',add.to.plot=TRUE, lty=2)
contour(k,add=T,nlevels=10)

image(k)
points(ca20, pch=20,cex.max=1,pt.divide='equal',add.to.plot=TRUE, lty=2)
contour(k,values=sqrt(k$krige.var),add=T,nlevels=10)
```

> Likelihood fit

```{r d25}
# Parametrisen mallin sovitus likfitin kautta
ml1st <- likfit(ca20, ini = c(80,20), 
                cov.model = "exp",
                fix.nugget=FALSE, trend="1st")
ml1st
plot(ca20.bin1st); lines(ml1st)

# universal kriging
krige.parml <- krige.control(type.krige='ok', 
                             cov.pars=ml1st$cov.pars, 
                             nugget=ml1st$nugget,
                             cov.model="exponential",
                             trend.d="1st", trend.l="1st")
kml <- krige.conv(ca20,locations=pred.grid,krige=krige.parml)
kml$beta.est

image(kml)
points(ca20, pch=20,cex.max=1,pt.divide='equal',add.to.plot=TRUE, lty=2)
contour(k, values=sqrt(k$krige.var), add=T, nlevels=10)
```

# 2-6 Random Fields Simulation

Conduct conditional simulation of `moreeni.dat` and compare the result with `RFsimulate` function.

```{r d26}
moreeni <- read.table("https://yan.fi/docs/sda/moreeni.dat",head=TRUE)
moreenigeo <- as.geodata(matrix(c(moreeni$x,moreeni$y,
                                  moreeni$pitoisuus),ncol=3))

library(RandomFields)
xgrid<-seq(min(moreeni$x),max(moreeni$x),l=100)
ygrid<-seq(min(moreeni$y),max(moreeni$y),l=100)
x <- runif(100, min=-1, max=1)
y <- runif(100, min=-1, max=1)
  #RandomFields Example
datatest <- RFsimulate(model = RMexp(), x=x, y=y, grid=FALSE)
  #simulate a field conditional on the above data
xyseq <- seq(-1.5, 1.5, length= 100)
cond <- RFsimulate(model = RMexp(), 
                   x = xyseq, y = xyseq, data=datatest)
plot(cond, datatest)

# our data
v2 <- variog(moreenigeo, estimator.type="classical", max.dist=90)
tul <- variofit(v2, fix.nug=FALSE,
                nugget=0.1, cov.model="gaussian",
                ini.cov.pars=c(1.5*10^7,60))
tul

cond <- RFsimulate(model=RMgauss(var=tul$cov.pars[1], scale=tul$cov.pars[2]), 
                   x=xgrid, y=ygrid, data=moreeni)
plot(cond, moreeni)
```




