---
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
    df_print: kable
    theme: 'readable'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(geoR)
```

# 1-4 Intro to GeoR

Check the data set `ca20` from `geoR`.

a) plot the data.

```{r d14a}
data(ca20)
plot(ca20$coords, cex=0.05*ca20$data,pch=16,col=gray(0.6))
plot.geodata(ca20)
```

b) is the data isotropic? Calculate directional semivariogram.

```{r d14b}
plot(variog4(ca20))
cat("extra: variance = sill = ", variog(ca20)$var.mark, var(ca20$data))
```

> "Not very likely isotropic. There exists trend.


c) suppose the data being isotropic. Calculate and explain the (empirical) semivariogram.

```{r d14c}
par(mfrow = c(1,2))
plot(variog(ca20))
plot(variog(ca20, max.dist=900))
varioc <- variog(ca20, max.dist=900) # classical
varioc$n
```

d) calculate robust semivariogram estimator and compare it with classic one.

```{r d14d}
varior <- variog(ca20, estimator.type="modulus", max.dist=900)
varior$n
par(mfrow=c(1,2))
plot(varior, ylim=range(varior$v), main = "robust")
plot(varioc, ylim=range(varior$v), main = "classic")
```

e) fit the classic semivarogram using multiple parameters and plot.

```{r d14e}
sovitegau <- variofit(varioc,
                      ini.cov.pars=c(150,600),
                      cov.model="gaussian",
                      fix.nugget=FALSE, nugget=50)
sovitegau

sovitegaunn <- variofit(varioc,
                        ini.cov.pars=c(150,600),
                        cov.model="gaussian",
                        fix.nugget=TRUE)
sovitegaunn

sovitesfa <- variofit(varioc,
                      ini.cov.pars=c(150,600),
                      cov.model="spherical",
                      fix.nugget=FALSE,nugget=50)
sovitesfa

plot(varioc,ylim=c(0,max(varior$v)), main = "classic")
lines(sovitegau, col = "blue") 
lines(sovitegaunn, col = "green")
lines(sovitesfa, col = "red")

sovitergau <- variofit(varior,
                       ini.cov.pars=c(150,600),
                       cov.model="gaussian",
                       fix.nugget=FALSE,nugget=50)
sovitergau

sovitersfa <- variofit(varior,
                       ini.cov.pars=c(150,600),
                       cov.model="spherical",
                       fix.nugget=FALSE,nugget=50)
sovitersfa

plot(varior,ylim=range(varior$v),main = "robust")
lines(sovitergau, col = "green")
lines(sovitersfa, col = "red")
```

> Cross-validation/Ristiinvalidointi

```{r d14ex}
# jai maku, etta kannattaisi sovittaa trendi!
xv.sovitegau <- xvalid(ca20, model = sovitegau)
mean(xv.sovitegau$std.error^2) 
par(mfcol = c(3,2)); par(mar = c(3,3,0,1)); par(mgp = c(2,1,0))
plot(xv.sovitegau,ask=FALSE,error=FALSE) 

xv.sovitergau <- xvalid(ca20, model = sovitergau)
mean(xv.sovitergau$std.error^2) 
par(mfcol = c(3,2)); par(mar = c(3,3,0,1)); par(mgp = c(2,1,0))
plot(xv.sovitergau,ask=FALSE,error=FALSE) 

xv.sovitesfa <- xvalid(ca20, model = sovitesfa)
mean(xv.sovitesfa$std.error^2) 
par(mfcol = c(3,2)); par(mar = c(3,3,0,1)); par(mgp = c(2,1,0))
plot(xv.sovitesfa,ask=FALSE,error=FALSE) 

xv.sovitesfa <- xvalid(ca20, model = sovitesfa)
mean(xv.sovitesfa$std.error^2) 
par(mfcol = c(3,2)); par(mar = c(3,3,0,1)); par(mgp = c(2,1,0))
plot(xv.sovitesfa,ask=FALSE,error=FALSE) 

xv.sovitersfa <- xvalid(ca20, model = sovitersfa)
mean(xv.sovitersfa$std.error^2) 
par(mfcol = c(3,2)); par(mar = c(3,3,0,1)); par(mgp = c(2,1,0))
plot(xv.sovitersfa,ask=FALSE,error=FALSE) 
```

> Klassiset sovittuisivat paremmin ja naista sfaarinen hieman paremmin (WLS pienempi ja mean(xv.sovitesfa$std.error^2) lahempana ykkosta)

f) Implement conventional kriging lattice with 25 $\times$ 25 resolution, and also include the error terms. Compare with the above.

```{r d14f}
# gaussisella mallilla kun nugget mukana?
ngridx<-25
ngridy<-25
xgrid<-seq(min(ca20$coords[,1]),max(ca20$coords[,1]),l=ngridx)
ygrid<-seq(min(ca20$coords[,2]),max(ca20$coords[,2]),l=ngridy)
pred.grid <- expand.grid(xgrid,ygrid)

krige.par <- krige.control(type.krige='ok',
                           cov.pars = sovitegau$cov.pars,
                           cov.model = sovitegau$cov.model, 
                           nugget = sovitegau$nugget)
k <- krige.conv(ca20,locations=pred.grid,krige=krige.par) 

plot(ca20$coords, cex=0.05*ca20$data,pch=16,col=gray(0.6))

image(k)
points(ca20, pch=20,cex.max=1,pt.divide='equal',add.to.plot=TRUE, lty=2)
contour(k,add=T,nlevels=10)

image(k)
points(ca20, pch=20,cex.max=1,pt.divide='equal',add.to.plot=TRUE, lty=2)
contour(k,values=sqrt(k$krige.var),add=T,nlevels=10)
```

```{r d14fx}
# sfaarisella mallilla kun nugget mukana?
krige.par <- krige.control(type.krige='ok',
                           cov.pars = sovitesfa$cov.pars,
                           cov.model = sovitesfa$cov.model, 
                           nugget = sovitesfa$nugget)
k <- krige.conv(ca20,locations=pred.grid,krige=krige.par) 

plot(ca20$coords, cex=0.05*ca20$data,pch=16,col=gray(0.6))

image(k)
points(ca20, pch=20,cex.max=1,pt.divide='equal',add.to.plot=TRUE, lty=2)
contour(k,add=T,nlevels=10)

image(k)
points(ca20, pch=20,cex.max=1,pt.divide='equal',add.to.plot=TRUE, lty=2)
contour(k,values=sqrt(k$krige.var),add=T,nlevels=10)

```