# Stochastic Simulation | Problems 5

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      comment = "") 
myprint <- function(printD, dig = 4)
  data.frame(round(printD, digits = dig)) 
```

##Exercise 1
<blockquote>
$p,q$ are densities, and $q(x)=0\Rightarrow p(x=0)$. Implement self-normalized importance sampling,

$$X_1,\dots,X_p\sim q(\cdot),\ \ W_k^{(n)}=\frac{w_u(X_k)}{\sum w_u(X_j)},\ \ w_u(x)=c_w\frac{p(x)}{q(x)}$$

a&b. Estimate $\text{Var}_p(X)$ and show that it is consistent.
</blockquote>

$$\begin{aligned} \text{Var}_p(X) &\longleftarrow 
 \sum \limits_{k=1}^{n} X_k^2 W_k^{(n)} - \bigg( \sum \limits_{k=1}^{n} X_k W_k^{(n)} \bigg)^2
\\ &= I_{p,q,MH}^{(n)}(f) - \bigg( I_{p,q,MH}^{(n)}(g) \bigg)^2
\\ &\overset{n\to\infty}\longrightarrow\ \mathbb E_p(X^2) - (\mathbb E_p(X))^2  &\text{the consistency of importance sampling}
\\ &= \text{Var}_p(X)
\\
\\ \text{where, } & f(x):=x^2 \text{ and }g(x):=x 
\end{aligned}$$

> Estimate quantiles and show the consistency

$$\begin{aligned} x_\beta &\longleftarrow \sum\limits_{k=1}^{n}\boldsymbol 1 (X_k\leq F_q^{-1}(\beta)) W_k^{(n)} & F_q^{-1} \text{ is the quantile of }q
\\ &= I_{p,q,MH}^{(n)}(h) &h(x):=\boldsymbol 1 (x\leq F_q^{-1}(\beta))
\end{aligned}$$

##Exercise 2 (not ok)
<blockquote>
Consider Metropolis-Hastings algorithm. Show the following

a. $(X_k, Y_{k+1})_{k\geq 1}$ is a Markov chain.

b. It admits $\tilde p(x,y)=p(x)q(x,y)$ as invariant distribution.
</blockquote>

For Metropolis-Hastings, $Y_{k+1}$ is dependent only on $X_k$, and $X_k$ only on $Y_k, X_{k-1}$ and a (non-relevant) i.i.d $U_k$. Therefore, the vector value $(X_k, Y_{k+1})$ is only conditional on $(X_{k-1}, Y_k)$, implying a proper Markov Chain.

Further, transition probability $P(X_{k},Y_{k+1} | X_{k-1},Y_{k})$ ...

##Exercise 3
<blockquote>
Metropolis-Hastings for $p_u([x,y])=\exp \bigg( -\frac 1 2 (x^2 + y^2) \bigg) + \exp \bigg( -\frac 1 2 ((x-7)^2 + y^2) \bigg)$ with proposal $q(x,y)=\tilde q(y-x)$ where $\tilde q \sim N(0, 4I_2)$.
</blockquote>

Background information/Notes

- the function `myMH.multi` only returns the simulated $X$ as matrix, while accepting initial values, log probability, proposal function, and also the function $f$.

- the initial value is purposefully chosen to be extreme in order to demostrate the burn-in

- in the proposal, the marginal distribution is used to simplify the proposal. This is proper due to the zero-correlation and properties of normal distribution

```{r demo53}
log_p_u<-function(x, m1=c(0,0), m2=c(7,0), S1=diag(2), S2=diag(2)) { 
  q1 <- -.5*t(x-m1) %*% S1 %*% (x-m1) 
  q2 <- -.5*t(x-m2) %*% S2 %*% (x-m2) 
  q0 <- max(q1, q2) # Avoid underflow of both exp’s below
  return(q0 + log(exp(q1-q0) + exp(q2-q0))) 
}

prop.multinorm <- function(x,y){
  # both 2-dim; since corr = 0, we can use the marginal
  return(c(x[1]+rnorm(1,0,2), x[2]+rnorm(1,0,2)))
}

myMH.multi <- function(n, burnIn, x, logpFun, propFun, fFun = function(x) x, debug = FALSE){
  totalN <- n + burnIn
  xs <- matrix(NA, length(x), totalN)
  accept <- 0
  l_px <- logpFun(x)
  
  for (i in 1:totalN) {
    # y from proposal
    y <- propFun(x)
    l_py <- logpFun(y)
    # acceptance?
    alpha <- exp(l_py - l_px)
    if(runif(1) < alpha){
      accept <- accept + 1 #set the acceptance ratio
      x <- y
      l_px <- l_py
    }
    xs[,i] <- fFun(x)
  }
  
  if(debug){plot(xs, type = "l");  print(xs)}
  
  return(xs[,-(1:burnIn)])
}

mySim <- myMH.multi(100000, 1000, c(0,9), 
                    logpFun = log_p_u, propFun = prop.multinorm)

library(coda)
eff1 <- effectiveSize(mySim[1,])
eff2 <- effectiveSize(mySim[2,])
iact1 <- 100000 / eff1
iact2 <- 100000 / eff2

par(mfrow = c(2,2))
plot(mySim[1,], type = "l"); mtext(paste("E[f1] =", mean(mySim[1,]), "\nVar[f1] = ", var(mySim[1,])))
acf(mySim[1,]); text(45, 0.8, paste("IACT", iact1, "\n n.eff", eff1));
plot(mySim[2,], type = "l"); mtext(paste("E[f2] =", mean(mySim[2,]), "\nVar[f2] = ", var(mySim[2,])))
acf(mySim[2,]); text(50, 0.8, paste("IACT", iact2, "\n n.eff", eff2));

varE1 <- var(mySim[1,]) / eff1
varE2 <- var(mySim[2,]) / eff2
data.frame(ci025 = c(f1 = mean(mySim[1,]) - 1.96 *varE1, f2 = mean(mySim[2,] - 1.96 * varE2)),
           ci975 = c(mean(mySim[1,]) + 1.96 *varE1, mean(mySim[2,] + 1.96 * varE2)))
```

##Exercise 4
<blockquote>
Consider the uniform distribution $p_a$ on $C_0\cup C_a$ where $C_t:=\{(x,y)\in\mathbb R^2: x \in [t,t+1],\ y \in [t,t+1]\}$
</blockquote>

It is two connecting squares (if $a\in[0,1]$), and the value of $a$ determines the second (up-right) square's location.

For the sampling, the idea is to sample from the "correct" rectangle. This task is solved mainly on the graph, instead of expressions.

The Gibbs sampler for $p_a$ starting with $x_0\in[0,a+1]$, and repeat with $k=1,2,\dots,n$:

(1) Sample $Y_k \sim U \bigg(t\cdot \boldsymbol 1 (X_{k-1} \gt 1),\ 1 + t\cdot\boldsymbol 1(X_{k-1}\geq t)\bigg)$

(2) Sample $X_k \sim U \bigg(t\cdot \boldsymbol 1 (Y_{k-1} \gt 1),\ 1 + t\cdot\boldsymbol 1(X_{Y-1}\geq t)\bigg)$

The sampler is not irreducible with $a=1$, but indeed irreducible for $a\in(0,1)$

##Exercise 5
<blockquote>
Continue from the previous task. Implement the Gibbs sampler for $p_{0.95}$ and calculate $E_{p_{0.95}}[f(X)],\ \ f(x,y)=x+y$
</blockquote>

```{r demo55b}
logp_uni <- function(x){
  # 3 rectangles -> uniform; otherwise, -infty
  if(x[1] > 0 && x[1] < 0.95 && x[2] > 0 && x[2] < 1)       return(0)
  if(x[1] > 0.95 && x[1] < 1 && x[2] > 0 && x[2] < 1.95)    return(0)
  if(x[1] > 1 && x[1] < 1.95 && x[2] > 0.95 && x[2] < 1.95) return(0)
  return(-Inf)
}


sim5b <- myMH.multi(1e5, 1000, c(0.5,0.5), 
                    logpFun = logp_uni, 
                    propFun = function(x) return(runif(2, 0, 1.95)),
                    fFun = function(x) return(sum(x)))
mean(sim5b)
```

> Implement the adaptive scaling Metropolis algorithm for $p_{0.95}$

```{r demo55c}
theta <-0; x <- c(0,0); n <- 1e5;  accept <- 0
xs <- matrix(NA, 2, n); thetas <- numeric(n);

l_px <- logp_uni(x)

for (k in 1:n) {
  y <- x + exp( theta) * runif(2)   #propose y
  l_py <- logp_uni(y)

  alpha <- min(1, exp(l_py - l_px)) #correct alpha with min
  
  if(runif(1) < alpha){
    accept <- accept + 1 #set the acceptance ratio
    x <- y
    l_px <- l_py
  }
  
  theta <- theta + k ^{-2/3} * (alpha - 0.234)
  xs[,k] <- x; thetas[k] <- theta;
}

cat('Ex = ', mean(xs[1,]), mean(xs[2,]), '\tEtheta = ', mean(thetas), '\tacceptance = ', accept/n)
```

The acceptance ration very low? Maybe because the adaptive scaling leading the simulation to the true value very quickly.
