# Yleistetty Lineaarinen Malli (Demo 6)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
load("ylm_0.RData")
```

## Exercise 1

Show that negative binomial is a distribution from exponential family

<blockquote>

$$\begin{aligned} f(y_i;\mu_i,\phi)
&= \exp \bigg( \log\frac{\Gamma(y_i+\phi)}{\Gamma(\phi)\Gamma(y_i+1)} + y_i \log \frac{\mu_i}{\mu_i + \phi} + \phi \log\frac{\phi} {\mu_i + \phi}\bigg)
\\ &= \exp \bigg(y_i \underbrace{\log \frac{\mu_i}{\mu_i + \phi}}_{:=\theta_i} + \underbrace{\phi \log \bigg( 1- e^{\frac{\mu_i}{\mu_i + \phi}}}_{:=b(\theta_i)=-\phi\log(1-e^{\theta_i})} \bigg) +  \underbrace{\log\frac{\Gamma(y_i+\phi)}{\Gamma(\phi)\Gamma(y_i+1)}}_{c(y_i,\phi)}\bigg)
\end{aligned}$$

</blockquote>

## Exercise 2

Fomulate $E(y_i)$ and $\text{Var}(y_i)$ for negative binomial

<blockquote>
$$\begin{aligned} E(y_i)
&= b'(\theta_i)
\\ &= \frac{\phi e^{\theta_i}}{1 - e^{\theta_i}}
\\ &= \bigg(\phi \frac{\mu_i}{\mu_i + \phi}  \bigg) \bigg/ \bigg( 1- \frac{\mu_i}{\mu_i + \phi} \bigg)
\\ &=\mu_i 
\\ \text{Var}(y_i) &= b''(\theta_i) a(\phi)
\\ & = \frac{d}{d\theta_i} \frac{\phi e^{\theta_i}}{1 - e^{\theta_i}}
\\ & = \frac{\phi e^{\theta_i}}{(1-e^{\theta_i})^2}
\\ & = \frac{\mu_i}{\frac {\phi}{\mu_i+\phi}}
\\ & = \frac{\mu_i (\mu_i +\phi)}{\phi}
\end{aligned}$$
</blockquote>

## Exercise 3R

Concerning butterfly data. 

<blockquote>
```{r d603}
options(stringsAsFactors = TRUE)

  # no site in model
fit603 <- glm(Colias ~ time + habitat + building, 
              family = poisson, 
              data = perhoset)
cbind(exp(coef(fit603)), exp(confint(fit603)))

b <- coef(fit603)
par(mfrow = c(2,2))
  #marginal plot on time
plot(Colias ~ time, pch = 18, col = "grey", 
     data = perhoset[perhoset$habitat == "Hayfield",], 
     main = "Yearly Change: Hayfield Habitat")
curve(exp(b[1] + b[2] * x + b[6] * mean(perhoset$building)), 
      from = 1, to = 5, add = TRUE)
plot(Colias ~ time, pch = 18, col = "grey",
     data = perhoset[perhoset$habitat == "Mixed",], 
     main = "Yearly Change: Mixed Habitat")
curve(exp(b[1] + b[2] * x  + b[3] + b[6] * mean(perhoset$building)), 
      from = 1, to = 5, add = TRUE)
plot(Colias ~ time,  pch = 18, col = "grey",
     data = perhoset[perhoset$habitat == "Short",], 
     main = "Yearly Change: Short Habitat")
curve(exp(b[1] + b[2] * x + b[4] + b[6] * mean(perhoset$building)), 
      from = 1, to = 5, add = TRUE)
plot(Colias ~ time,  pch = 18, col = "grey",
     data = perhoset[perhoset$habitat == "Tall",], 
     main = "Yearly Change: Tall Habitat")
curve(exp(b[1] + b[2] * x + b[5] + b[6] * mean(perhoset$building)), 
      from = 1, to = 5, add = TRUE)

  # marginal plot on building
plot(Colias ~ building, pch = 18, col = "grey", 
     data = perhoset[perhoset$habitat == "Hayfield",], 
     main = "Effect of Building Density: Hayfield Habitat")
curve(exp(b[1] + b[2] * 3 + b[6] * x), 
      from = 0, to = 20, add = TRUE)
plot(Colias ~ building, pch = 18, col = "grey",
     data = perhoset[perhoset$habitat == "Mixed",], 
     main = "Effect of Building Density: Mixed Habitat")
curve(exp(b[1] + b[2] * 3  + b[3] + b[6] * x), 
      from = 0, to = 20, add = TRUE)
plot(Colias ~ building,  pch = 18, col = "grey",
     data = perhoset[perhoset$habitat == "Short",], 
     main = "Effect of Building Density: Short Habitat")
curve(exp(b[1] + b[2] * 3 + b[4] + b[6] * x), 
      from = 0, to = 20, add = TRUE)
plot(Colias ~ building,  pch = 18, col = "grey",
     data = perhoset[perhoset$habitat == "Tall",], 
     main = "Effect of Building Density: Tall Habitat")
curve(exp(b[1] + b[2] * 3 + b[5] + b[6] * x), 
      from = 0, to = 20, add = TRUE)
```
</blockquote>

## Exercise 4R

Continue from above, formulate the risk ratio.

<blockquote>
The table above, together with confidence interval was indeed already in risk ratio.
</blockquote>

## Exercise 5R

Continue from above, use the property tool to test the hypothesis that different type of habitats have the same expected amount

<blockquote>
Since the types of habitats are categrical (factors), and the "hidden" information is that the first factor (`Hayfield`) equals 0; therefore, the null hypothesis can be simplified to $\beta_3=\beta_4 = \beta_5 = 0$. The F-test will be used by fixing $\boldsymbol K^T = \boldsymbol I_3$ and $\boldsymbol m = (0\ 0\ 0)^T$.

Part of code is referenced from the previous exercises.

```{r d605}
F_testi <- function(K, X, beta, m, sigma)
{
  q <- qr(K)$rank; n <- nrow(X); p <- ncol(X)
  C <- solve(t(K) %*% solve(t(X) %*% X) %*% K)
  F <- t(t(K) %*% beta - m) %*% C %*% (t(K) %*% beta - m) / (q * sigma)
  p_arvo <- 1 - pf(F, q, n - p - 1)
  list(F = F, p_arvo = p_arvo)
}

K <- diag(nrow = 3)
m <- matrix(rep(0,3), nrow = 3)
X <- data.frame(x1 = as.numeric(perhoset$habitat == "Mixed"),
                x2 = as.numeric(perhoset$habitat == "Short"),
                x3 = as.numeric(perhoset$habitat == "Tall"))
X <- as.matrix(X)
sigma2 <- t(residuals(fit603)) %*% residuals(fit603) / (nrow(X) - length(b) -1)
beta <- matrix(b[3:5], nrow = 3)

F_testi(K, X, beta, m, sqrt(sigma2))
```

Reject null hypothesis, i.e. habitats make difference.

</blockquote>

## Exercise 6R

With the same dataset, try negative-binomial.

<blockquote>
```{r d606}
library(MASS)
fit606 <- glm.nb(Colias ~ time + habitat + building, data = perhoset)
cbind(exp(coef(fit606)), exp(confint(fit606)))
```

The result is very similar with Poisson-regression
</blockquote>

## Exercise 7R

Concerning circulation disease data, using poisson model to analyze the death rate among different genders

<blockquote>
```{r d607}

fit607 <- glm(kuolleet ~ vuosi * sukup + offset(log(vaki)), 
              family = poisson, data = vk)
rr <- cbind(exp(coef(fit607)), exp(confint(fit607)))
prob <- rr / (rr + 1)
prob[-1,]

b <- coef(fit607)
vk$lk <- vk$kuolleet / vk$vaki * 100000
plot(lk ~ vuosi, data = vk[vk$sukup == 1,], col = "red", ylim = c(25,350),
     main = "circulation disease death rate, with fit")
points(lk ~vuosi, data = vk[vk$sukup == 0,], col = "blue")
curve(exp(b[1] + b[2] * x + b[3] + b[4] * x) * 1e5, col = "red",
      from = 1970, to = 2014, add = TRUE)
curve(exp(b[1] + b[2] * x ) * 1e5, col = "blue",
      from = 1970, to = 2014, add = TRUE)
legend("topright", c("man","woman"), col = c("blue","red"), pch = 1)
```
</blockquote>

## Exercise 8R

Continue from above, estimate the death rate for the year 2015-2020 (in genders) together with the confidence interval

<blockquote>
```{r d608, rows.print = 12, full.width = T}
new <- data.frame(vuosi = c(2015:2020, 2015:2020),
                  sukup = c(rep(0,6), rep(1,6)),
                  vaki = 1e5)
fitted <- predict(fit607, newdata = new, type = "response", se.fit = TRUE)
data.frame(vuosi = new$vuosi,
           sukup = new$sukup,
           est = fitted$fit,
           lwr = fitted$fit - 1.96 * fitted$se.fit,
           upr = fitted$fit + 1.96 * fitted$se.fit)
```
</blockquote>